D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-06 ootavacuous.tex
Add implementation-constraints section
While in the area, upgrade to the most recent C++ standard, remove
all references to compilers, and start the rework of the section
looking at OOTA definitions.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{abstract}
	The out-of-thin-air (OOTA) properties of the specification
	of \co{memory_order_relaxed} have resulted in considerable
	consternation over the years.
	Attempts to create memory models that rule out OOTA behaviors
	have either been non-executable, complex, or unloved by compiler
	writers.
	But at the same time, there are no known instances of OOTA
	behavior in real C++ implementations.

@@

All OOTA workers owe a debt to the foundational work in the infamous
``Causality Test Cases''.\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by compiler implementers.

@@

We could follow this long-established tradition of proposing
solutions to the OOTA problem without a definition of this problem.
However, we (perhaps foolishly) felt the need to provide a clear
definition, which is a more detailed version of the wording in 33.5.4p8
(\co{[atomics.order]}):

@@

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work, and by the laws of physics, which have
not.
Additional constraints are imposed by hardware architecture and
design, which have been partially accounted for in the standard,
one limitation being that accurate and executable formal descriptions
of hardware memory models did not appear until after the standard
was released.

The following sections discuss these constraints, starting with the
laws of physics, continuing with hardware constraints, and ending
with constraints imposed by the standard.

@@

As a special case of this prohibitino, a value stored speculatively must
not be exposed to non-speculative loads from other threads.
In the further special case of multiple hardware threads sharing a core,
one might imagine speculation involving multiple threads.
However, in this case, any squashing of speculation must squash the
full extent of that speculation, across all of the hardware threads
that were involved.

@@

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2022N4910},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

The reason that atomic stores cannot be invented is that doing so can
introduce new (and almost certainly undesirable) behaviors that are
forbidden by the abstract machine.
To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
a CPPMEM\footnote{
	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
litmus test that demonstrates such a behavior.
Without line~4, only the values~2 and~3 can be stored to \co{r1}.
With that line, the additional value~42 can also be stored to \co{r1}.
The compiler is therefore forbidded from inventing that store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
For example, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

Duplicating atomic stores can also introduce new and undesireable
behaviors.
To see this, consider Listing~\ref{lst:CPPMEM Duplicated Store}, a CPPMEM
litmus test that demonstrates such a behavior.
Without line~4, if the final value of \co{r1} is 1, then the final value
of \co{x} must be 2.
With that line, the final value of \co{x} can be 1 even when the
final value of \co{r1} is 1.
The compiler is therefore forbidded from duplicating that store of 1
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
Again, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

\paragraph{Omitted Redundant Atomic Stores}
In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
might always be executed such that no other thread accesses \co{x}
during the time between those two stores.
This means that, if the implementation acted as if the store was omitted
from the source, the user would have no way to prove that fact short of
inspecting the assembly code.
In cases where such omissions are undesirable, the user can resort to
volatile atomic stores or to inline assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent the compiler from omitting a store.
However, acting as if such a redundant store was omitted from the source
code cannot create an OOTA cycle because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of non-volatile
relaxed atomic stores omitted is essentially a valid execution of the
program with the store present.

@@

The concept of flattening multiple threads into a single thread is
tantalizing, especially given the large cache-miss penalties inherent
to large multicore systems, which can have many sockets containing
thousands of CPUs.
One might object to the whole concept of flattening as a compiler
optimization given the difficulty inherent in avoiding deadlocks
and other hazards.
Nevertheless, it is only reasonable to ask whether flattening can
result in OOTA cycles.
As we will see, the answer is ``no''.

@@

Note that if both legs of the \co{if} statement spanning lines~7-10,
the compiler would be within its rights to hoist the store out of
that \co{if} statement.
However, in this case, the two stores on lines~8 and~10 are storing
distinct values, preventing the compiler from hoisting.

@@

Therefore, if your C++ implementation produces OOTA cycles, you have
bugs that would be considered bugs even if OOTA cycles were allowed.
And you need to fix those bugs, regardless of whether they are in your
compiler, your hardware, or your alternative universe.

@@

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
However, the \co{herd7} tool does not find this litmus test's cycle, presumably
because this tool does not do the deep inter-thread compiler analysis
required to find it.

>>>>

\begin{abstract}
	The out-of-thin-air (OOTA) properties of the specification
	of \co{memory_order_relaxed} have resulted in considerable
	consternation over the years.
	Attempts to create memory models that rule out OOTA behaviors
	have either been non-executable, complex, or unloved by C++
	implementers.
	But at the same time, there are no known instances of OOTA
	behavior in real C++ implementations.

@@

All OOTA workers owe a debt to the foundational work in the infamous
``Causality Test Cases'',\footnote{
	\url{http://www.cs.umd.edu/~pugh/java/memoryModel/unifiedProposal/testcases.html}.}
which may be found in
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}.

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.

@@

The C++ standard says this in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:

@@

Others, current authors included, have define an OOTA cycle as a
cycle in sdep $\cup$ rf, where sdep is the set of semantic
dependencies within each thread and rf is the set of store-to-load
links, whether intra-thread or between threads.
This is a fine definition, and is consistent with the words in the C++
standard, except that it simply shunts all the complexity into the
term ``semantic dependency''.

We could follow this long-established tradition of proposing
solutions to the OOTA problem without a definition of this problem.
However, we (perhaps foolishly) felt the need to provide a clear
definition,

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work, and by the laws of physics, which have
not.
Additional constraints are imposed by hardware architecture and
design, which have been partially accounted for in the standard,
one limitation being that accurate and executable formal descriptions
of hardware memory models did not appear until after the standard
was released.
A final set of constraints is imposed by the C++ implementations
themselves.

The following sections discuss these constraints, starting with the
laws of physics, continuing with hardware constraints, continuing
further with constraints imposed by the standard, and culminating
with constraints imposed by the C++ implementations themselves.

@@

As a special case of this prohibition, a value stored speculatively must
not be exposed to non-speculative loads from other threads.
In the further special case of multiple hardware threads sharing a core,
one might imagine speculation involving multiple threads.
However, in this case, any squashing of speculation must squash the
full extent of that speculation, across all of the hardware threads
that were involved.

@@

Because volatile atomics constitute observable
behavior~\cite[\co{intro.abstract}]{ThomasKoeppe2023N4950},
they must be executed in strict accordance with the rules of the abstract
machine.
This means that any volatile atomic operation involving a store
must execute as if all loads whose return values are used to compute
that store's address or value have already returned.
OOTA afficienados will recognize this as a special case of ordering
relaxed loads before relaxed stores, albeit one not requiring
expensive memory-fence instructions on weakly ordered architectures.

@@

The reason that atomic stores cannot be invented is that doing so can
introduce new (and almost certainly undesirable) behaviors that are
forbidden by the abstract machine.
To see this, consider Listing~\ref{lst:CPPMEM Invented Store},
a CPPMEM\footnote{
	\url{http://svr-pes20-cppmem.cl.cam.ac.uk/cppmem/index.html}.}
litmus test that demonstrates such a behavior.
Without line~4, only the values~2 and~3 can be stored to \co{r1}.
With that line, the additional value~42 can also be stored to \co{r1}.
The implementation is therefore forbidded from inventing that store of 42
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
For example, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

@@

Duplicating atomic stores can also introduce new and undesireable
behaviors.
To see this, consider Listing~\ref{lst:CPPMEM Duplicated Store}, a CPPMEM
litmus test that demonstrates such a behavior.
Without line~4, if the final value of \co{r1} is 1, then the final value
of \co{x} must be 2.
With that line, the final value of \co{x} can be 1 even when the
final value of \co{r1} is 1.
The implementation is therefore forbidden from duplicating that store of 1
unless it can prove that doing so does not negatively affect the
program's observable behaviors, independently of OOTA cycles.
Again, the implementation might be able to prove that there
are no other accesses to \co{x} at the time of the invented store.

\paragraph{Omitted Redundant Atomic Stores}
In contrast, a pair of back-to-back non-volatile atomic stores to \co{x}
might always be executed such that no other thread accesses \co{x}
during the time between those two stores.
This means that, if the implementation acted as if the store was omitted
from the source, the user would have no way to prove that fact short of
inspecting the assembly code.
In cases where such omissions are undesirable, the user can resort to
volatile atomic stores or to inline assembly\footnote{
	For example, by placing the Linux-kernel \co{barrier()} macro
	between the two stores.
	This macro is an empty GCC \co{asm} that specifies the \co{memory}
	clobber.}
to prevent the implementation from omitting a store.
However, acting as if such a redundant store was omitted from the source
code cannot create an OOTA cycle because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of non-volatile
relaxed atomic stores omitted is essentially a valid execution of the
program with the store present.

\subsection{C++ Implementations}
\label{sec:C++ Implementations}

C++ implementations execute C++ programs, and use a range of techniques.
For example, many implementations take steps to optimize the code to
be executed.
This optimization presents a trade-off between the time spent optimizing
the code and the time spent actually executing it.
One type of optimization is the identification of non-semantic dependencies
and the replacement of them with simpler computations, up to and including
replacement of arbitrarily complex passages of code with constants.\footnote{
	There are a great many other optimizations, but this paper
	necessarily focuses on semantic dependencies.}

This optimization process may, roughly speaking, be divided into two
activities, analysis of the code and classification of dependencies.

\subsubsection{Code Analysis}
\label{sec:Code Analysis}

As would be expected given the various optimization/execution tradeoffs,
different C++ implementations carry out different degrees of analysis.

First, at one extreme, ``simple implementations'' do absolutely no
analysis beyond that absolutely required to correctly execute the program.

Second, other implementations do local analysis.
This analysis might be on a per-code-path basis, within a single function,
within a single function after inlining, and so forth.

Third, one could imagine implementations whose analysis is restricted to
a thread.
In this case, the analysis stops at loads to objects that were last
stored to by other threads (``rfe''), stores to objects where the next
and/or previous stores in that object's modification order were executed
by other threads (``coe'')), and loads from objects for which the value
returned by a given load was overwritten by some other thread (``fre'').

Fourth, one could also imagine implementations whose analysis is
completely unrestricted, up to and including the limiting case where
execution is oracular in nature.
Of course, if we are going to imagine omniscient analysis and oracles,
we can also rely on them to locate any OOTA cycles that might appear
in environments not constrained by the laws of physics, hardware
limitations, and C++ implementation concerns.

However, for more realistic implementations in the third category,
analysis relevant to OOTA cycles begins with loads from objects that were
last stored to by some other thread and ends with stores to objects that
might be loaded by other threads.

If additional inter-thread information is available, the corresponding
constraints may be applied.
For example, if an implementation determines that the values of a pair
of objects are always equal at any time that a given thread might access
them, one could imagine a sufficiently brave implementation substituting
a load from one for a load from the other.

The next section focuses on the key analysis step for OOTA cycles,
namely dependency classification.

\subsubsection{Dependency Classification}
\label{sec:Dependency Classification}

Carrying out this type of optimization requires that the implementation
classify dependencies as semantic or not.
Given the limited resources available, almost all implementations will
choose to approximate sdep, which entails some misclassifications, with
the consequences summarized in
Table~\ref{tab:Semantic-Dependency Classification}.
The first row of the table depicts the case where the implementation
correctly identifies a semantic dependency, and thus exhibits temporal
behavior for the corresponding code.
The second row depicts the case where the implementation misclassifies
the dependency as non-semantic, which will result in an incorrect
execution.
In this case, the proper course of action is to fix that bug.
The third row depicts the case where the implementation misclassifies
the dependency as semantic, thus missing a chance to optimize.
The resulting execution is nevertheless correct, and is also temporal.
The fourth and final row depicts the case where the implementation
correctly classifies the dependency as non-semantic, in which case
the execution is both correct and atemporal.

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Code	& Implementation	& Result \\
\hline
sdep		& sdep			& Temporal \\
\cline{2-3}
		& $\neg$sdep		& BUG!!! \\
\hline
$\neg$sdep	& sdep			& Temporal \\
\cline{2-3}
		& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification}
\label{tab:Semantic-Dependency Classification}
\end{table}

Note well that in the absence of bugs in the implementation (which
again should be fixed), it is the implementation's classification of
the dependency that dictates whether or not the corresponding execution
is temporal.

Therefore, in the limiting case, analyses of OOTA cycles that focus on a
given implementation can simply use that implementation's classification
of dependencies as semantic on the one hand or non-semantic on the other.
While taking this approach sacrifices significant generality, it has the
benefit of greatly reduces the cost of dependency analysis, potentially
to zero.

@@

The concept of flattening multiple threads into a single thread is
tantalizing, especially given the large cache-miss penalties inherent
to large multicore systems, which can have many sockets containing
thousands of CPUs.
One might object to the whole concept of flattening as an
optimization given the difficulty inherent in avoiding deadlocks
and other hazards.
Nevertheless, it is only reasonable to ask whether flattening can
result in OOTA cycles.
As we will see, the answer is ``no''.

@@

Note that if both legs of the \co{if} statement spanning lines~7-10,
the implementation would be within its rights to hoist the store out of
that \co{if} statement.
However, in this case, the two stores on lines~8 and~10 are storing
distinct values, preventing the hoisting.

@@

Therefore, if your C++ implementation produces OOTA cycles, you have
bugs that would be considered bugs even if OOTA cycles were allowed.
And you need to fix those bugs, regardless of whether they are in your
C++ implementation, your hardware, or your alternative universe.

@@

Listing~\ref{lst:Causality Test Case 1}
shows causality test case 1, for which the \co{r1 == r2 == 1} result
is to be allowed.
However, the \co{herd7} tool does not find this litmus test's cycle, presumably
because this tool does not do the deep inter-thread program analysis
required to find it.
